---
title: "blood_vs_cortex-analysis"
author: "Daniel Gregory"
format: html
editor: visual
---

# Intro

[Broad institute code](https://github.com/broadinstitute/gtex-tutorials/blob/master/GTEx_data_tutorial_R.ipynb) was used as basis for analysis.

# Sample & donor metadata

Load GTEx annotation files.

```{r}
sample.df  <- read.delim("data/GTEx_Analysis_v11_Annotations_SampleAttributesDS.txt",
                         as.is = TRUE, header = TRUE, row.names = 1)
subject.df <- read.delim("data/GTEx_Analysis_v11_Annotations_SubjectPhenotypesDS.txt",
                         as.is = TRUE, header = TRUE, row.names = 1)
```

Visualize GTEx annotation files. Note:

- SUBJID -\> unique subject ID

- SEX -\> 1 = male, 2 = female

- DTHHARDY = Hardy scale -\> 0 = ventilator, 1 = violent and fast (\< 10 min), 2 = fast of natural causes (\< 1 hr), 3 = intermediate (1 - 24 hours), 4 = slow (\> 1 day)

```{r}
head(sample.df[0:3,])
head(subject.df[0:3,])

cat("Sample dimensions: ", dim(sample.df), "\n")
cat("Subject dimensions: ", dim(subject.df), "\n")
```

Build a new data frame containing RNA-seq analysis freeze samples only. Appears that samples that *aren't* included in GTEx are initially included in the sample attributes annotations.

```{r}
rnaseq.sample.df <- sample.df[sample.df['SMAFRZE']=='RNASEQ',]

head(rnaseq.sample.df[1:5,])
```

Within the GTEx sample annotations, the number of samples within each tissue site can be visualized

```{r}
#plot(sort(table(rnaseq.sample.df['SMTSD']), decreasing=TRUE))
```

Prior to bring in the expression data, we can isolate relevant annotation information for our target tissue sites.

```{r}
blood.sample.df = rnaseq.sample.df[rnaseq.sample.df['SMTSD'] == 'Whole Blood', ]
cortex.sample.df = rnaseq.sample.df[rnaseq.sample.df['SMTSD'] == 'Brain - Cortex', ]
```

# Import expression data

Remove .gct two-line header. data.table will be faster than delim for the large file size.

```{r}
library(data.table)
blood.dt <- fread("data/gene_reads_adult_gtex_v11_whole_blood.gct", skip=2)
cortex.dt <- fread("data/gene_reads_adult_gtex_v11_brain_cortex.gct", skip=2)
```

Data info:

```{r}
head(blood.dt[,1:6]) # When rendered, all rows will be printed -> limit to 6
head(cortex.dt[,1:6])

cat("Whole-blood dimensions: ", dim(blood.dt), "\n")
cat("Brain-cortex dimensions: ", dim(cortex.dt), "\n")
```
