---
title: "Brain stress score by Hardy scale"
author: "Daniel Gregory"
format: html
editor: visual
---

# Intro

Investigating how brain stress is affected by Hardy scale.

Gene set:

- **Glucocorticoid (stress hormone) response:** FKBP5, TSC22D3, ZBTB16, KLF9 ([Dal Molin et al. 2026, Mol Neurobiol](https://pmc.ncbi.nlm.nih.gov/articles/PMC13341972/)).

# Settings

```{r}
gc_genes <- c("FKBP5", "TSC22D3", "ZBTB16", "KLF9")
```

# Load data

```{r}
library(data.table)
library(DESeq2)

sa <- fread("data/raw/GTEx_Analysis_v11_Annotations_SampleAttributesDS.txt")[SMAFRZE == "RNASEQ"]
sa[, SUBJID := sapply(strsplit(SAMPID, "-"), function(p) paste(p[1], p[2], sep = "-"))]
subj <- fread("data/raw/GTEx_Analysis_v11_Annotations_SubjectPhenotypesDS.txt")

# Read one tissue's counts, normalize within that tissue, return log2 expression
load_tissue <- function(file, tissue) {
  dt <- fread(file, skip = 2)
  samples <- sa[SMTSD == tissue & SAMPID %in% names(dt)]
  counts <- as.matrix(dt[, samples$SAMPID, with = FALSE])
  rownames(counts) <- dt$Description # gene symbols
  norm <- t(t(counts) / estimateSizeFactorsForMatrix(counts))
  list(expr = log2(norm + 1), samples = samples)
}

cortex <- load_tissue("data/raw/gene_reads_adult_gtex_v11_brain_cortex.gct", "Brain - Cortex")
```

# Scores

Score = the average expression (log2 normalized counts) of the gene set in that tissue.

```{r}
score <- function(t, genes) setNames(colMeans(t$expr[match(genes, rownames(t$expr)), ]), t$samples$SUBJID)
setdiff(gc_genes, rownames(cortex$expr)) # gene symbols not found (should be empty)

d <- data.table(SUBJID = cortex$samples$SUBJID)
d[, brain_stress := score(cortex, gc_genes)[SUBJID]]
d <- subj[d, on = "SUBJID"] # donor info

nrow(d) # donors with cortex
```

# How donors died affects the stress score

Hardy scale: 0 = ventilator, 1 = violent/fast, 2 = fast natural, 3 = intermediate, 4 = slow.

```{r}
table(d$DTHHRDY) # donors per group
boxplot(brain_stress ~ DTHHRDY, data = d, xlab = "Hardy scale", ylab = "Brain stress score")
```
