---
title: "DESeq2 differential expression"
author: "Daniel Gregory"
format: html
editor: visual
---

# Intro

[Duke HTS 2017 DESeq2 introduction](https://people.duke.edu/~ccc14/duke-hts-2017/Statistics/08032017/DESeq2-Notebook-introduction.html) = base for this analysis. Minors changes were made for the more recent DESeq2 release and adapting to a different dataset.

Input is from the tissue_overlap-analysis.qmd at present.

# Settings

Program is input-blind outside of these settings.

```{r}
input_file <- "data/processed/deseq_input_blood_vs_cortex.rds" # "data/processed/deseq_input_blood_vs_sigmoid.rds"

sample_col <- "SAMPID" # coldata column holding sample IDs (= count matrix column names)
group_col <- "tissue" # coldata column that defines the comparison
target_level <- "Brain - Cortex" # upregulated "Colon - Sigmoid"
ref_level <- "Whole Blood" # downregulated = reference level

block_col <- NULL # "SUBJID" = paired design. NULL = unpaired comparison
covariates <- NULL # extra design terms - unpaired only. c("SEX", "AGE", "DTHHRDY")

# I'll have to explore these to find the best values
alpha <- 0.05 # FDR cutoff for calling DE genes
lfc_cutoff <- 1.5 # |log2FC| used to highlight hits in the volcano plot
min_count <- 10 # prefilter -> reads a sample needs for a gene to count as expressed in it
min_frac <- 0.5 # prefilter -> keep genes expressed in >= this fraction of samples in either group
shrink_type <- "apeglm" # lfcShrink method: "apeglm" (recommended), "ashr", or "normal"

out_dir <- "results"
```

Clean input values. DESeq2 doesn't accept spaces or dashes.

```{r}
clean <- function(x) gsub("[^A-Za-z0-9]+", "_", x)
target <- clean(target_level)
ref <- clean(ref_level)

comparison <- paste(target_level, "vs", ref_level)
file_tag <- paste0(target, "_vs_", ref)

# group goes last so results()/resultsNames() default to the comparison of interest
design_formula <- reformulate(c(block_col, covariates, "group"))
design_formula
```

# Load packages

```{r}
library(DESeq2)
library(ggplot2)
```

# Import data

The upstream file saved a list with the count matrix, the sample table (colData), and gene ID -\> symbol annotation.

```{r}
inp <- readRDS(input_file)
genes <- inp$genes # Name = Ensembl ID, Description = gene symbol

# Keep only the two levels being compared
coldata <- inp$coldata
coldata <- coldata[coldata[[group_col]] %in% c(target_level, ref_level), ]

# Build the design columns - reference level first
coldata$group <- factor(clean(coldata[[group_col]]), levels = c(ref, target))
for (v in c(block_col, covariates)) {
  coldata[[v]] <- factor(clean(as.character(coldata[[v]])))
}

# DESeq2 can't use samples with missing design values. DTHHRDY can have some NAs
design_vars <- c(block_col, covariates, "group")
has_na <- !complete.cases(coldata[, design_vars, drop = FALSE])
if (any(has_na)) message("Dropping ", sum(has_na), " samples with missing design values")
coldata <- coldata[!has_na, ]

rownames(coldata) <- coldata[[sample_col]]
coldata[[sample_col]] <- NULL # redundant with the rownames
counts <- inp$counts[, rownames(coldata)]

table(coldata$group)
```

Import into DESeqDataSet. Upstream data = count matrix so use FromMatrix in place of fromHTSeqCount.

```{r}
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = coldata,
                              design = design_formula)
```

## Inspect object

```{r}
dds
class(dds)
slotNames(dds)
```

Accessor functions to inspect data.

```{r}
head(colData(dds))
design(dds)
```

Raw counts: genes along the rows, samples along the columns.

```{r}
dim(counts(dds))
head(counts(dds)[, 1:6], 3)
```

## Prefilter low-count genes

DESeq2 vignette recommends dropping genes with very small amounts of reads to accelerate computation.

The vignette's rule (\>= min_count reads in \>= smallest-group-size samples) is too strict with GTEx. Instead consider the groups individually

```{r}
keep <- sapply(levels(dds$group), function(g) {
  in_g <- dds$group == g
  rowSums(counts(dds)[, in_g] >= min_count) >= min_frac * sum(in_g)
})
keep <- rowSums(keep) > 0
table(keep)
dds <- dds[keep, ]
```

# Fit the model

`DESeq()` runs the full pipeline: size factors -\> dispersions -\> Wald test. Also replaces outlier counts and refits those genes when sample size is \>= 7.

```{r}
dds <- DESeq(dds)
dds # sizeFactor added to colData; baseMean, dispersion, Wald results, ... added to rowData
resultsNames(dds)[length(resultsNames(dds))] # Print
```

## Size factors

Each sample is given a size factor.

```{r}
round(head(sizeFactors(dds)), 3)
summary(sizeFactors(dds))
```

Quick aside to view differences in size factors between tissue samples.

```{r}
boxplot(sizeFactors(dds) ~ dds$group, xlab = NULL, ylab = "Size factor")
```

Normalized counts are raw counts divided by the size factors.

```{r}
head(counts(dds)[, 1:6], 3)
head(counts(dds, normalized = TRUE)[, 1:6], 3)
```

## Dispersion parameters

The fitted mean-dispersion trend.

```{r}
dispersionFunction(dds)
```

One dispersion per gene.

```{r}
alphas <- dispersions(dds)
length(alphas)
round(alphas[1:5], 4)
mcols(dds)[1:4, c("baseMean", "baseVar", "dispGeneEst", "dispFit", "dispersion", "dispOutlier")]
summary(alphas)
boxplot(log(alphas), ylab = "log(dispersion)")
```

Plot creation from vignette. Final should fit tighter to the line than the gene estimate.

With large sample sizes this effect would likely be less pronounced and a poor fit wouldn't necessarily be cause for concern. Unclear at this time what the threshold is though ,"as the sample size increases, the shrinkage decreases in strength, and eventually becomes negligible" [(Love et. al, 2014)](https://pmc.ncbi.nlm.nih.gov/articles/PMC4302049/). The vignette has example has only 7 samples.

Shrinkage hasn't been run yet either - explored later in the code.

```{r}
plotDispEsts(dds)
```

# Differential expression analysis

The Wald test already ran inside `DESeq()` above (calculates log2FoldChange, lfcSE, stat, pvalue).

Haven't run a "zero-centered Normal prior (betaPrior)" akaa shrinkage yet.

The alpha threshold should be redundant because the of min_counts filtering.

```{r}
res <- results(dds, contrast = c("group", target, ref), alpha = alpha)
res[1:4, ]
data.frame(desc = mcols(res)$description)
```

## P-values

Get unadjusted and adjusted p values. The adjusted p-values have "independent filtering" automatically applied by DESeq so they may differ slightly from hand calculations.

```{r}
pvalues <- res$pvalue
adjp <- res$padj
pvalues[1:4]
adjp[1:4]
```

## Summarize, sort, subset

```{r}
summary(res)
```

Attach gene symbols, then sort by p-value.

```{r}
res$symbol <- genes$Description[match(rownames(res), genes$Name)]
res_ord <- res[order(res$pvalue), ]
head(res_ord)
```

## Volcano plot

Many padj values underflow to exactly 0, so -log10(padj) = Inf and plot() drops those genes. Cap the y-axis instead and draw capped genes as triangles along the top. Small p-values are discussed in this [thread](https://support.bioconductor.org/p/117706/).

```{r}
# Original - drops genes with padj = 0
# plot(res$log2FoldChange,-log10(res$padj),pch=19,cex=0.3,xlab="Log2 FC",ylab="-log10(BH Adjusted P-value)",col=ifelse(res$padj<0.05&abs(res$log2FoldChange)>lfc_cutoff,"red","black"))

y <- -log10(res$padj)
y_cap <- 300
capped <- !is.na(y) & y > y_cap
y[capped] <- y_cap
hit <- res$padj < alpha & abs(res$log2FoldChange) > lfc_cutoff
# red = higher in target_level (right), blue = higher in ref_level (left)
plot(res$log2FoldChange, y, pch = ifelse(capped, 17, 19), cex = 0.3,
     xlab = paste("Log2 FC", target_level, "vs", ref_level), ylab = "-log10(BH adjusted p-value), capped at 300",
     col = ifelse(!hit, "black", ifelse(res$log2FoldChange > 0, "red", "blue")))
```

## Shrunken log2 fold changes

`DESeq()` no longer shrinks fold changes, it's now done separately within `lfcShrink`. Shrinking is generally advised to improve ranking and visualization - doesn't change p-values.

```{r}
if (shrink_type != "normal" && !requireNamespace(shrink_type, quietly = TRUE)) {
  message(shrink_type, " is not installed; falling back to type = 'normal'. ",
          "Install with BiocManager::install('", shrink_type, "')")
  shrink_type <- "normal"
}

res_shrunk <- lfcShrink(dds, coef = paste0("group_", target, "_vs_", ref),
                        type = shrink_type, svalue = shrink_type != "normal") # s-values need apeglm or ashr
res_shrunk$symbol <- res$symbol
head(res_shrunk[order(res$pvalue), ]) # svalue output has no pvalue column; rows match res
```

MA plots before and after shrinkage. Noisy low-count genes get pulled toward zero. The unshrunken estimate is capped by DESeq2's `minmu = 0.5` floor, while apeglm refits from raw counts without it.

```{r}
DESeq2::plotMA(res, alpha = alpha, main = "Unshrunken")
DESeq2::plotMA(res_shrunk, alpha = alpha, main = paste("Shrunken (", shrink_type, ")"))
```

## Save results

```{r}
dir.create(file.path(out_dir, "deseq2"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(out_dir, "shinygo", "input"), recursive = TRUE, showWarnings = FALSE)
out <- as.data.frame(res_ord)
out$log2FoldChange_shrunk <- res_shrunk[rownames(out), "log2FoldChange"]
out <- cbind(gene_id = rownames(out), out)
write.csv(out, file.path(out_dir, "deseq2", paste0("deseq2_", file_tag, ".csv")), row.names = FALSE)
```

## Gene lists for ShinyGO

Directional hit lists plus the background (every gene that was tested). Ensembl IDs with the version suffix stripped ("ENSG00000302013.1" -\> "ENSG00000302013"), one per line, ready to paste into ShinyGO.

```{r}
strip_version <- function(x) unique(sub("\\..*$", "", x))
sig <- !is.na(res$padj) & res$padj < alpha

gene_lists <- list(
  up = strip_version(rownames(res)[sig & res$log2FoldChange > lfc_cutoff]), # higher in target
  down = strip_version(rownames(res)[sig & res$log2FoldChange < -lfc_cutoff]), # higher in reference
  background = strip_version(rownames(res))
)
names(gene_lists) <- c(paste0("up_in_", target), paste0("up_in_", ref), "background")

for (n in names(gene_lists)) {
  writeLines(gene_lists[[n]], file.path(out_dir, "shinygo", "input", paste0("shinygo_", file_tag, "_", n, ".txt")))
}
lengths(gene_lists)
```

## Save output for figures

Results table, normalized counts, and sample info in one file.

```{r}
res_out <- as.data.frame(res)
res_out$gene_id <- rownames(res)
saveRDS(list(res = res_out,
             norm_counts = counts(dds, normalized = TRUE),
             coldata = as.data.frame(colData(dds)),
             settings = list(target_level = target_level, ref_level = ref_level, target = target, ref = ref,
                             alpha = alpha, lfc_cutoff = lfc_cutoff, comparison = comparison, file_tag = file_tag)),
        file.path("data", "processed", paste0("deseq2_output_", file_tag, ".rds")))
```

Fitted DESeq2 object.

```{r}
saveRDS(dds, file.path("data", "processed", paste0("deseq2_dds_", file_tag, ".rds")))
```

# Transformations for clustering and PCA

## Variance stabilizing transformation

Transformation to remove dependence of the variance on the mean. `vst()` will be faster than `rlog()` for 100s of samples as it avoid fitting a shrinkage term.

**`blind = true`** forces the re-estimation of the dispersion "using only an intercept." This ensures there is no bias between the experimental groups. —\> Can cause fittings problems if there are large differences in counts (which is likely the case here). Bias will be fairly small so worth exploring both options.

```{r}
vsd <- vst(dds, blind = TRUE)
class(vsd)
VSTmat <- assay(vsd)
dim(VSTmat)
VSTmat[1:10, 1:6]
```

Hierarchical clustering.

```{r}
dists <- dist(t(VSTmat))
hc <- hclust(dists)
plot(hc, labels = FALSE, xlab = NULL, sub = NULL)
```

PCA.

```{r}
plotPCA(vsd, intgroup = "group")
```

# Session information

```{r}
sessionInfo()
```
