---
title: "tissue_overlap-analysis"
author: "Daniel Gregory"
format: html
editor: visual
---

# Intro

[Broad institute code](https://github.com/broadinstitute/gtex-tutorials/blob/master/GTEx_data_tutorial_R.ipynb) was used as basis for analysis.

# Settings

Select tissues and files for analysis.

```{r}
tissue_a <- "Whole Blood" # tissue names from the SMTSD column
tissue_b <- "Brain - Cortex" # "Colon - Sigmoid"

file_a <- "data/raw/gene_reads_adult_gtex_v11_whole_blood.gct"
file_b <- "data/raw/gene_reads_adult_gtex_v11_brain_cortex.gct" # "data/raw/gene_reads_adult_gtex_v11_colon_sigmoid.gct"

# Save under data folder to avoid uploading to git
output_file <- "data/processed/deseq_input_blood_vs_cortex.rds" # "data/processed/deseq_input_blood_vs_sigmoid.rds"
```

# Sample & donor metadata

Load GTEx annotation files.

```{r}
library(data.table)
sample.dt  <- fread("data/raw/GTEx_Analysis_v11_Annotations_SampleAttributesDS.txt")
subject.dt <- fread("data/raw/GTEx_Analysis_v11_Annotations_SubjectPhenotypesDS.txt")
```

Visualize GTEx annotation files. Note:

- SUBJID -\> unique subject ID

- SEX -\> 1 = male, 2 = female

- AGE -\> Decade

- DTHHARDY = Hardy scale -\> 0 = ventilator, 1 = violent and fast (\< 10 min), 2 = fast of natural causes (\< 1 hr), 3 = intermediate (1 - 24 hours), 4 = slow (\> 1 day)

```{r}
head(sample.dt, 5)
head(subject.dt, 5)

cat("Sample dimensions: ", dim(sample.dt), "\n")
cat("Subject dimensions: ", dim(subject.dt), "\n")
```

Build a new data table containing RNA-seq analysis freeze samples only. Appears that samples that *aren't* included in tissue expression data are initially included in the sample attributes annotations.

```{r}
rnaseq.sample.dt <- sample.dt[SMAFRZE == "RNASEQ"]

head(rnaseq.sample.dt, 5)
```

Within the GTEx sample annotations, the number of samples within each tissue site can be visualized

```{r}
# matrix(sort(table(rnaseq.sample.dt$SMTSD), decreasing=TRUE))
```

Prior to bringing in the expression data, we can isolate relevant annotation information for our target tissue sites.

```{r}
a.sample.dt = rnaseq.sample.dt[SMTSD == tissue_a]
b.sample.dt = rnaseq.sample.dt[SMTSD == tissue_b]
```

# Import expression data

Remove .gct two-line header.

Optional: move description (gene ID) to gene variable to allow querying - might not be needed for data tables.

```{r}
a.dt <- fread(file_a, skip=2)
b.dt <- fread(file_b, skip=2)
```

Data info (note description row is included in genes now):

```{r}
head(a.dt[,1:6]) # When rendered, all rows will be printed -> limit to 6
head(b.dt[,1:6])

cat(tissue_a, "dimensions: ", dim(a.dt), "\n")
cat(tissue_b, "dimensions: ", dim(b.dt), "\n")
```

# Cohort overlap

Build overlap and isolation datasets.

```{r}
# Subject annotation uses reduced donor ID
# Build function to remove non-uniform ID components
#   "GTEX-1117F-0005-SM-HL9SH"  ->  "GTEX-1117F"
donor_id <- function(x) {
  parts <- tstrsplit(x, "-", fixed = TRUE)  # split each ID into its fields
  paste(parts[[1]], parts[[2]], sep = "-")  # rejoin the first two = the donor
}

# Isolate donor IDs in each group
a.donors <- donor_id(a.sample.dt$SAMPID)
b.donors <- donor_id(b.sample.dt$SAMPID)

# Split the donors into relevant groups
overlap.donors <- intersect(a.donors, b.donors)  # gave both tissues
a.only.donors  <- setdiff(a.donors,  b.donors)   # gave tissue a, not b
b.only.donors <- setdiff(b.donors, a.donors)    # gave tissue b, not a

# Show sizes of each group
data.table(
  group   = c(tissue_a, tissue_b, "Overlap (both)",
              paste(tissue_a, "only"), paste(tissue_b, "only")),
  n_donors = c(length(a.donors), length(b.donors),
               length(overlap.donors), length(a.only.donors),
               length(b.only.donors))
)

# Make venn diagram
library(eulerr)
region.counts <- c(length(a.only.donors), length(b.only.donors), length(overlap.donors))
names(region.counts) <- c(tissue_a, tissue_b, paste(tissue_a, tissue_b, sep = "&"))

plot(euler(region.counts), quantities = TRUE, main = "Donors by tissue")
```

## Demographics

Covariate data: sex (1 = male, 2 = female), age (sorted by 10s), hardy scale.

Compare overlap with original sample distributions.

I reviewed the sample attributes and didn't see anything that would be relevant for selection bias control.

```{r}
# Label every donor with its group, then attach that donor's phenotypes.
# Create list w/ donor IDs and labeled group column x5 -> stack vertically with rbindlist
donor.groups <- rbindlist(list(
  data.table(group = "overlap",    SUBJID = overlap.donors),
  # data.table(group = paste(tissue_b, "only"), SUBJID = b.only.donors),
  # data.table(group = paste(tissue_a, "only"),  SUBJID = a.only.donors),
  data.table(group = tissue_b, SUBJID = b.donors),
  data.table(group = tissue_a, SUBJID = a.donors)
))

# Attach covariate data from subject dataset
donor.groups <- subject.dt[donor.groups, on = "SUBJID"] #data.table join using key to add demographic info


# For each covariate -> print counts + normalized table -> stacked bar chart
library(ggplot2)
for (i in c("SEX", "AGE", "DTHHRDY")) {

  # Pull out the two columns we care about for this covariate
  work <- data.table(
    covariate = as.character(donor.groups[[i]]),  # the covariate value
    group = donor.groups$group                # which cohort the donor is in
  )

  # Count donors in each group-level combination (.N = rows in each bucket)
  counts <- work[, .(n = .N), by = .(group, covariate)]

  # Turn each count into a proportion of its group
  counts[, group_total := sum(n), by = group]     # total donors in the group
  counts[, prop := round(n / group_total, 3)]     # count / group total = share of group

  # Print counts + proportions side by side -> dcast pivots the long list into a wide grid
  cat("\n---", i, "---\n")
  print(dcast(counts, covariate ~ group, value.var = c("n", "prop"), fill = 0))

  # Plot
  p <- ggplot(counts, aes(x = group, y = prop, fill = covariate)) +
    geom_col(width = 0.75) +
    labs(x = NULL, y = "Proportion within group", fill = i,
         title = paste("Composition by group -", i))
  print(p)
}

```

## Paired-donor subset for a matched comparison

```{r}
# Grab the paired samples from each tissue
a.paired  <- a.sample.dt [donor_id(SAMPID) %in% overlap.donors]
b.paired <- b.sample.dt[donor_id(SAMPID) %in% overlap.donors]

# Stack into one table
paired.dt <- rbind(a.paired, b.paired)
paired.dt <- paired.dt[, .(SAMPID, SUBJID = donor_id(SAMPID), tissue = SMTSD)]

# Attach donor covariates
paired.dt <- subject.dt[paired.dt, on = "SUBJID"]

# Set column order + sort so each donor's two rows sit together
setcolorder(paired.dt, c("SAMPID", "SUBJID", "tissue", "SEX", "AGE", "DTHHRDY"))
setorder(paired.dt, SUBJID, tissue)

paired.dt
```

## Expression counts for the paired samples

```{r}
# The paired sample IDs for each tissue
a.samp  <- paired.dt[tissue == tissue_a, SAMPID]
b.samp <- paired.dt[tissue == tissue_b, SAMPID]

# Both files share the same gene rows -> column-bind
counts <- cbind(as.matrix(a.dt [, ..a.samp]),
                as.matrix(b.dt[, ..b.samp]))
rownames(counts) <- a.dt$Name # label rows with gene IDs

# Put columns in the same order as the colData rows
counts <- counts[, paired.dt$SAMPID]

dim(counts)
counts[1:5, 1:4]
```

## Save DESeq2 inputs

Bundle the count matrix, sample table, and gene annotation into one file for the analysis file.

```{r}
# Sanity checks: same gene order in both files, colData rows match count columns
stopifnot(identical(a.dt$Name, b.dt$Name),
          identical(colnames(counts), paired.dt$SAMPID))

saveRDS(list(counts = counts,
             coldata = as.data.frame(paired.dt),
             genes = as.data.frame(a.dt[, .(Name, Description)])),
        output_file)
```
